---
title: "Discrete-Time Model of Seashell Population Recovery from Industrial Pollution"
author: "Melo Filho; Zhu Qingdong"
date: "2024-10-16"
output:
  html_document:
    df_print: paged
---

Initially, there is a population of seashells at one point in the river. Downstream from this point, there is an entry of contaminated effluents into the river. Further downstream, after the entry of these effluents, there is a spot in the river where it would be possible for a population of the same species of seashells to stabilise. However, due to the pollution, the number of individuals in this area remains consistently low.

Next, we want to consider a situation of substantial migration from point A, resulting in the arrival of individuals with mutations that would enable them to survive the existing pollution.

With this, we aim to explore how much migration would be necessary to reach a tipping point, where the resistant population could establish itself and grow, despite the presence of pollution.



***

## Simulating evolutionary rescue in discrete time

In this notebook, we are implementing simulations of evolutionary rescue by a single mutation in discrete time, as discussed and analyzed in the work by Orr & Unckless. The model is as follows: a haploid population starts with a given number of wild-type and mutant individuals. Because of a sudden environmental change, the wild type is unfit and declines in the new environment. The mutant is beneficial and can survive and expand in the new environment, but it starts at a low (or zero) frequency in the total population. In order to rescue the population, it has to outcompete the wild type. If the mutant is not present at time 0, it has to appear through mutation and then spread in the population before extinction occurs.

## List of variables and parameters

- $N_a$ current size of the population "a"  (changes over time)
- $N_mut$ current size of the mutant (A) who migrated from the population "a" (changes over time)
- $N_b$ current size of the population "b" (changes over time)
- $T$ Time
- N_migra_rate precentage of population who migrate from point A to piont B in population A
- N_init_a initial population size of population "A" (should be a positive integer)
- N_init_b initial population size of population "B"  (should be a non-negative integer)
- decay_rate detrimental effect of the environment on the population "B" (i.e., the wild-type population decays at this rate; should be >0 to induce potential extinction)
- t_max number of generations after which to stop the simulation (should be an integer >0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 1.5 times the original total population size, so t_max can be set to a large value without creating extremely long simulation times.




## Simulate one generation of population size change
We want to observe the numbers of mutant of population A who surviver the migration . As first step, we simulate one generation of the population A. At time $T$, the population B has $N_b$ current size of the population "b" and $N_ma$ current size of the mutant (A) who migrated from the population "a". First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.
```{r}
simulate_one_gen <- function(N_a, N_bw, N_bm, pollution_rate,sel_coeff, migra_rate, mut_rate){
  # draw offspring_a according to Poisson distribution
  offsp_a <- rpois(1, N_a * (1))
  # draw new migration according to Poisson distribution
  migra_a_to_b <- rpois(1, offsp_a * migra_rate)
  # draw offspring_b according to Poisson distribution
  offsp_bw <- rpois(1, N_bw * (1-pollution_rate))
  offsp_bm <- rpois(1, ((N_bw+migra_a_to_b) * mut_rate)*(1+sel_coeff))
  # determine new population sizes of population B plus migration
  # N_a_new <- max((offsp_a - migra_a_to_b) * 1.25, 0)
  N_b_new <- max(floor(offsp_bw+offsp_bm+migra_a_to_b), 0)
  # draw new mutants according to Poisson distribution 
  N_mut <- rpois(1, (N_b_new) * mut_rate)
  return(c(N_mut, N_b_new))
}
# Test the function
print(simulate_one_gen(N_a =1000, N_bw =1000, N_bm = 0, pollution_rate =.7, sel_coeff=.2, migra_rate =.2, mut_rate =.01))
``` 



### Questions
* Why do we substract `mut_a_to_A` from `N_a_new` and not from `N_A_new`?
* In the life cycle as defined above, do individuals reproduce first and then mutate, or vice versa?
* Running the test output repeatedly, why do you almost always get mutant individuals in the next generation, even though you start without any mutants?
* What does the function `c` do?

## Simulate a population trajectory
We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.
```{r}
set.seed(5)
simulate_pop <- function(N_bw, N_bm, pollution_rate, sel_coeff, migra_rate, mut_rate, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_bw, N_bm)
  # initiate the variables
  pop_new <- c(N_bw, N_bm)
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2], pollution_rate, sel_coeff, migra_rate, mut_rate, t_max)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
    # condition to stop the simulation before t_max: either the population exceeds 1.5 times the original population size, or it goes extinct
    if (pop_new[1]+pop_new[2]>=1.5*(N_bw+N_bm) | pop_new[1]+pop_new[2]==0) break
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("bw","bm")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 1000
# create your simulation data
output <- simulate_pop(1000, 0, 0.7, 0.2, 0.2, 0.01, max_gen)
# show the last few lines of the data table
print(tail(output))
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")
``` 

### Questions
* What is the difference between `pop_vector` and `pop_new`? 
* What is the condition that breaks the loop good for? What are its disadvantages?
* What does the function `tail` do?

## A small analysis
As you can see if you simulate and plot the output several times, there is a lot of stochasticity/variation in the outcome. Therefore, to get interpretable results, we have to run the simulation many times and summarize the outcome depending on the question of our study. Here I am showing an example in which we record the time and the value of the lowest population size (or the time of extinction, if the population size eventually goes to 0). Usually, we want to study this for a range of model parameters; here I am varying the decay rate and the selection coefficient. Other (maybe more interesting) features to study would be the following:

- rescue probability dependent on initial frequency and selection coefficient
- time at which the rescue mutation occurs first if it rescues, vs. if it doesn't rescue, dependent on initial population size and mutation rate
- time until the population has recovered in case it becomes rescued, dependent on mutation rate and selection coefficient
- ...

```{r}
# set some parameters to fixed values
init_a <- 100
init_A <- 0
m_rate <- 0.001
max_gen <- 1000
# determine how often to run the simulation for each set of parameters
no_replicates <- 100

# set parameters to vary
s_values <- c(0.2,0.3,0.4)
r_values <- c(0.05,0.1,0.15)

# initialize data table - where to collect the results
data_table <- c()
# run the simulation across all chosen parameters
# loop over decay rates
for(rval in r_values){
  # loop over selection coefficients
  for(sval in s_values){
    # different way of running many simulations: make replicates using "repeat" function with a counter i
    # reset counter
    i<-1
    repeat {
      # increase counter by one
      i<-i+1
      # run the simulation once
      one_run <- simulate_pop(init_a,init_A,rval,sval,m_rate,max_gen)
      # determine total population sizes
      total_size <- one_run[,1]+one_run[,2]
      # determine minimum population size
      min_size <- min(total_size)
      # determine (first) generation at which this population size occurred
      min_gen <- as.numeric(which(total_size==min_size)[1])
      # enter the data into the table
      data_table <- rbind(data_table,c(rval,sval,min_gen,min_size)) # note that we add the varying parameters (decay rate and selection coefficient) to the table too
      # stop the repeated computation after no_replicates times
      if(i>no_replicates) break
    }
  }
}
# define column names of data table
colnames(data_table) <- c("r","s","min_gen","no_min")
# show the first lines of the output
print(head(data_table))
``` 

### Questions
* Why does this code block run much more slowly than the others?
* Why is it useful to define the parameters at the beginning (e.g. `init_a=100`) and use their names rather than the numbers throughout the rest of the code?
* What does the function `which` do?

## Making your results reproducible
When you generate “random numbers” in R, you are actually generating pseudorandom numbers. These numbers are generated with an algorithm that requires a seed to initialize. This means that if you know the seed and the generator, you can reproduce the output.

```{r}
# Draw 5 (pseudo)random numbers from a normal distribution
rnorm(5)
# Draw again 5 (pseudo)random numbers from a normal distribution
rnorm(5)
``` 

Expectedly, the numbers are different. How can we reproduce the exact output?

```{r}
#initialize the seed of the pseudo random generator
set.seed(1)
# Draw 5 (pseudo)random numbers from a normal distribution
rnorm(5)
#initialize the seed of the pseudo random generator
set.seed(1)
# Draw again 5 (pseudo)random numbers from a normal distribution
rnorm(5)
``` 

## Next steps
So, where do you go from here? Of course you won't just stare at these tables (at least not forever). Visualization would be a good step to see how the minimum population size varies with the decay rate and the selection coefficient. And you might want to include only the cases in which rescue happens, and also extract the rescue probability in parallel. However, the idea is that you pick a question that you find (more) interesting and implement modified versions of this code to address your question of choice.
