---
title: "Basic model of evolutionary rescue in discrete time with immigration and competition"
author: "Melo Filho; Zhu Qingdong"
date: "2024-10-25"
output:
  html_document:
    df_print: paged
---

In this notebook, we explore a modified model of evolutionary rescue, introducing stochastic immigration and competition through logistic growth. The primary biological question addressed is: *"In the event of an increase in the number of migrants, what is the estimated timeframe for reaching carrying capacity, and how might the decay rate affect this outcome?"*

------------------------------------------------------------------------

## Simulating evolutionary rescue in discrete time

In this notebook, we implement simulations of the evolutionary rescue of a mussel population in a river, aided by migration from an upstream population and a single beneficial mutation in discrete time, as discussed and analyzed in the work by Orr & Unckless. The model is as follows: a population begins with a given number of wild-type individuals. Due to a sudden environmental change, caused by pollution reaching this area, the wild type becomes unfit and declines in the new environment. A mutant variant, however, is beneficial and can survive and expand in the polluted environment, though it starts at a low (or zero) frequency in the population. In order to rescue the population, this mutant must outcompete the wild type. If the mutant is not present at time 0, it must appear through mutation and spread within the population before extinction occurs.

## List of variables and parameters

-   $N_a$ current size of the wild-type (a) population (changes over time)
-   $N_A$ current size of the mutant (A) population (changes over time)
-   $N_total$ current total population size, representing the sum of wild-type and mutant individuals.
-   `carrying_capacity` maximum sustainable population size; the population growth rate declines as this limit is approached due to environmental constraints.
-   `N_init_a` initial population size of wild-type population (should be a positive integer)
-   `N_init_A` initial population size of the mutant population (should be a non-negative integer)
-   `decay_rate` detrimental effect of the environment on the wild type (i.e., the wild-type population decays at this rate; should be \>0 to induce potential extinction)
-   `sel_coeff` selective advantage of the mutant over the wild type (should be \>decay_rate to allow for evolutionary rescue)
-   `mut_rate` mutation rate from the wild type to the mutant genotype (should be \>0 and small)
-   `mig_rate_a` is the immigration rate of "a" individuals
-   `t_max` number of generations after which to stop the simulation (should be an integer \>0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 1.5 times the original total population size, so `t_max` can be set to a large value without creating extremely long simulation times.

## Simulate one generation of population size change

We want to observe the numbers of wild-type and mutant individuals over time. As first step, we simulate one generation of the population. At time $t$, the population has $N_a$ wild-type individuals and $N_A$ mutant individuals. First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.

## Include competition with logistic growth and migration

Steps of the life cycle:

*adult migrants of type* $a$ arrive reproduction: $a$ individuals always suffer from the decay rate, independent of the current population size; the growth rate of $A$ individuals decreases to 1 as the carrying capacity is approached \*mutation occurs

```{r}


# The following function simulates one generation, incorporating immigration and competition effects. Wild-type individuals are subject to decay, while mutant growth decreases as carrying capacity is approached.

simulate_one_gen_mig_logistic <- function(N_a, N_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a) {
  # adult migrants arrive and are added to 'a' population
  N_a_mig <- rpois(1, mig_rate_a) # stochastic immigration
  N_a <- N_a + N_a_mig
  
  # determine total population size
  N_total <- N_a+N_A
  
  # reproduction occurs according to decay rates and selection coefficients
  offsp_a <- rpois(1, N_a * (1-decay_rate)) # independent of carrying capacity
  offsp_A <- rpois(1, N_A + N_A*(sel_coeff-decay_rate)*(1-N_total/carrying_capacity))
  # randomly kill individuals if carrying capacity is exceeded
  offsp_total <- offsp_A + offsp_a
  if(offsp_total > carrying_capacity)
  {
    offsp_a <- rbinom(1,carrying_capacity,c(offsp_a,offsp_A)/offsp_total)
    offsp_A <- carrying_capacity-offsp_a
  }

  
  # draw new mutants according to Poisson distribution
  mut_a_to_A <- rpois(1, offsp_a * mut_rate)
  
  # determine new population sizes of wild type and mutant
  N_a_new <- offsp_a - mut_a_to_A
  N_A_new <- offsp_A + mut_a_to_A
  
  return(c(N_a_new, N_A_new))
}
# Test the function
print(simulate_one_gen_mig_logistic(N_a=50, N_A=0, carrying_capacity=50, decay_rate=0.1, sel_coeff=0.3, mut_rate=0.1, mig_rate_a=1))

```

## Simulate a population trajectory

Tracking population dynamics over generations helps us observe changes, but for a complete picture, we must follow the population from time 0 until either the mutant type rescues it or it goes extinct. The function below simulates this process, capturing the dynamics between wild-type and mutant populations across generations.

```{r}
simulate_pop_mig_logistic <- function(N_init_a, N_init_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_mig_logistic(pop_new[1],pop_new[2],carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("a","A")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 100
# create your simulation data
output <- simulate_pop_mig_logistic(1000,0,1000,0.1,0.3,0.001,3,max_gen)
# show the last few lines of the data table
print(tail(output))
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")
```

## What could be interesting analysis outputs from this model?

Important to note: extinction is impossible because new migrants always come in. But it seems that for high immigration rates and relatively low carrying capacity, the mutant can repeatedly get lost even if it is advantageous in the environment. This complicates the analysis.

### What could we measure?

-   Time until the carrying capacity is reached. *This indicates that the mutant has taken over and "rescued" the population -- at least temporarily.*
-   Average frequency of the two types over the last XXX (number to be determined, e.g. 100 or 1000?) generations of the simulation. *This describes how much the population is suffering from the inflow of deleterious genotypes.*
-   Count whether or how often or how fast the mutant is lost again after "rescuing" the population. *This describes how sensitive the rescued population is to swamping from the upstream population.*

### Which parameters to vary, and which to keep constant?

## Find time to carrying capacity

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

```{r}
simulate_pop_mig_logistic <- function(N_init_a, N_init_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a, t_max) {
  
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  # start counting generations
  gen <- 0
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_mig_logistic(pop_new[1],pop_new[2],carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a)
    # adding one generation to counter
    gen <- gen+1
    # calculate population size
    N_total <- sum(pop_new)
    # stop when carrying capacity is reached
    if (N_total == carrying_capacity) break
  }
  return(gen)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 100
# create your simulation data
output_gens <- simulate_pop_mig_logistic(1000,0,1000,0.1,0.3,0.001,3,max_gen)
print(output)
```
##what is the chunk？
This chunk is designed to test the number distribution of the two genotypes in each generation
### What is the difference between the simulate_pop_mig_logistic and simulate_pop_mig_logistic_2？
simulate_pop_mig_logistic_2 can show the number of different genotype in pop_new.simulate_pop_mig_logistic can only show the total number

```{r}
simulate_pop_mig_logistic_2 <- function(N_init_a, N_init_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a, t_max) {
  
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  # start counting generations
  gen <- 0
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_mig_logistic(pop_new[1],pop_new[2],carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a)
    # adding one generation to counter
    gen <- gen+1
    # calculate population size
    N_total <- sum(pop_new)
    # stop when carrying capacity is reached
    if (N_total == carrying_capacity) break
  }
  return(pop_new[2])	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 100
# create your simulation data
output_gens <- simulate_pop_mig_logistic(1000,0,1000,0.1,0.3,0.001,3,max_gen)
print(output)
```
### What could we measure?

-   testing the time to carrying capacity when all the parameters are constant
```{r}

N_init_a <-  1000
N_init_A <- 0
carrying_capacity <- 1000
decay_rate <- 0.1
sel_coeff <- 0.3
mut_rate <- 0.001
mig_rate_a <- 3
t_max <- 1000
n_reps <- 500


table_matrix <- matrix(nrow = n_reps, ncol = 1) 
print(table_matrix)  


# repeat the simulation n_reps times

 for (rep in 1:n_reps) {
   
# start the simulation with the same initials conditions
table_matrix[rep,1] <- simulate_pop_mig_logistic(N_init_a, N_init_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a, t_max)}
 
  print(table_matrix)  

  ## Calculate the average to carrying capacity
  avg_Na <- mean(table_matrix)
  #print(avg_NA)
  



```

### What could we measure?

-   Trying to make the change of time to carrying capacity with the change of the mean number of migrants be visible.

```{r}
N_init_a <- 1000
N_init_A <- 0
carrying_capacity <- 1000
decay_rate_a <- 0.1
decay_rate_b <- 0.15
decay_rate_c <- 0.2
sel_coeff <- 0.3
mut_rate_values <- c(0.001, 0.005)  
mig_rate_a <- seq(1, 200, by=2)
t_max <- 1000
n_reps <- 1000

# Table to store results for different mutation rates and decay rates
table_avg_a_001 <- matrix(nrow = length(mig_rate_a), ncol= 2)
table_avg_b_001 <- matrix(nrow = length(mig_rate_a), ncol= 2)
table_avg_c_001 <- matrix(nrow = length(mig_rate_a), ncol= 2)

table_avg_a_005 <- matrix(nrow = length(mig_rate_a), ncol= 2)
table_avg_b_005 <- matrix(nrow = length(mig_rate_a), ncol= 2)
table_avg_c_005 <- matrix(nrow = length(mig_rate_a), ncol= 2)

# Loop over each mutation rate
for(mut_rate in mut_rate_values){
  
  q <- 1
  for(mig in mig_rate_a){
    
    # Start the simulation for different decay rates (A, B, C)
    table_matrix <- matrix(nrow = n_reps, ncol= 1)
    
    # For decay_rate_a
    for (rep in 1:n_reps) {
      table_matrix[rep, 1] <- simulate_pop_mig_logistic(N_init_a, N_init_A, carrying_capacity, decay_rate_a, sel_coeff, mut_rate, mig, t_max)
    }
    avg_gen_time <- mean(table_matrix)
    if(mut_rate == 0.001){
      table_avg_a_001[q, 1] <- mig
      table_avg_a_001[q, 2] <- avg_gen_time
    } else {
      table_avg_a_005[q, 1] <- mig
      table_avg_a_005[q, 2] <- avg_gen_time
    }
    
    # For decay_rate_b
    table_matrix <- matrix(nrow = n_reps, ncol= 1)
    for (rep in 1:n_reps) {
      table_matrix[rep, 1] <- simulate_pop_mig_logistic(N_init_a, N_init_A, carrying_capacity, decay_rate_b, sel_coeff, mut_rate, mig, t_max)
    }
    avg_gen_time <- mean(table_matrix)
    if(mut_rate == 0.001){
      table_avg_b_001[q, 1] <- mig
      table_avg_b_001[q, 2] <- avg_gen_time
    } else {
      table_avg_b_005[q, 1] <- mig
      table_avg_b_005[q, 2] <- avg_gen_time
    }
    
    # For decay_rate_c
    table_matrix <- matrix(nrow = n_reps, ncol= 1)
    for (rep in 1:n_reps) {
      table_matrix[rep, 1] <- simulate_pop_mig_logistic(N_init_a, N_init_A, carrying_capacity, decay_rate_c, sel_coeff, mut_rate, mig, t_max)
    }
    avg_gen_time <- mean(table_matrix)
    if(mut_rate == 0.001){
      table_avg_c_001[q, 1] <- mig
      table_avg_c_001[q, 2] <- avg_gen_time
    } else {
      table_avg_c_005[q, 1] <- mig
      table_avg_c_005[q, 2] <- avg_gen_time
    }
    
    q <- q + 1
  }
}



# Plot the results for both mutation rates
plot(table_avg_a_001[, 1], table_avg_a_001[, 2], type = 'l', col = "blue", ylim = c(0, max(
  table_avg_a_001[, 2], table_avg_b_001[, 2], table_avg_c_001[, 2], 
  table_avg_a_005[, 2], table_avg_b_005[, 2], table_avg_c_005[, 2]
)), xlab = "Mean number of migrants", ylab = "Generations to Carrying Capacity", main = "Comparison of Decay Rates for Different Mutation Rates")

# Add lines for mutation rate = 0.001 (blue, red, purple)
lines(table_avg_b_001[, 1], table_avg_b_001[, 2], col = "red")
lines(table_avg_c_001[, 1], table_avg_c_001[, 2], col = "purple")

# Add lines for mutation rate = 0.005 (green, orange, yellow)
lines(table_avg_a_005[, 1], table_avg_a_005[, 2], col = "green")
lines(table_avg_b_005[, 1], table_avg_b_005[, 2], col = "orange")
lines(table_avg_c_005[, 1], table_avg_c_005[, 2], col = "yellow")


```


### What could we measure?

-   In the process which is mentioned by last chuck，we try to classify which genotype dominant when the mean number of migrants



```{r}

N_init_a <- 1000
N_init_A <- 0
carrying_capacity <- 1000
decay_rate_a <- 0.1
decay_rate_b <- 0.15
decay_rate_c <- 0.2
sel_coeff <- 0.3
mut_rate <- 0.005
mig_rate_a <- seq(1, 200, by=2)
t_max <- 1000
n_reps <- 500

# Table to store results for decay_rate = 0.1
table_avg_a <- matrix(nrow = length(mig_rate_a), ncol= 2)

# Table to store results for decay_rate = 0.15
table_avg_b <- matrix(nrow = length(mig_rate_a), ncol= 2)

# Table to store results for decay_rate = 0.2
table_avg_c <- matrix(nrow = length(mig_rate_a), ncol= 2)


q <- 1
#  increase migration
for(mig in mig_rate_a){

 # repeat the simulation n_reps times
  table_matrix <- matrix(nrow = n_reps, ncol= 1  )

 for (rep in 1:n_reps) {
   
# start the simulation with the same initials conditions with the decay rate A
table_matrix[rep,1] <- simulate_pop_mig_logistic_2(N_init_a, N_init_A, carrying_capacity, decay_rate_a, sel_coeff, mut_rate, mig, t_max)
}

  ## Calculate the average to carrying capacity
  avg_gen_time <- mean(table_matrix)
  table_avg_a[q, 1] <- mig
  table_avg_a[q,2] <- avg_gen_time
  
  q <- q+1
  
}
  
 
# start the simulation with the same initials conditions with the decay rate B

q <- 1
#  increase migration
for (mig in mig_rate_a) {

 # repeat the simulation n_reps times
  table_matrix <- matrix(nrow = n_reps, ncol = 1)

 for (rep in 1:n_reps) {
   
# start the simulation with the same initials conditions with the decay rate B
table_matrix[rep,1] <- simulate_pop_mig_logistic_2(N_init_a, N_init_A, carrying_capacity, decay_rate_b, sel_coeff, mut_rate, mig, t_max)
}

  ## Calculate the average to carrying capacity
  avg_gen_time <- mean(table_matrix)
  table_avg_b[q, 1] <- mig
  table_avg_b[q,2] <- avg_gen_time
  
  q <- q+1
  
}  
# start the simulation with the same initials conditions with the decay rate B

q <- 1
#  increase migration
for (mig in mig_rate_a) {

 # repeat the simulation n_reps times
  table_matrix <- matrix(nrow = n_reps, ncol = 1)

 for (rep in 1:n_reps) {
   
# start the simulation with the same initials conditions with the decay rate C

table_matrix[rep,1] <- simulate_pop_mig_logistic_2(N_init_a, N_init_A, carrying_capacity, decay_rate_c, sel_coeff, mut_rate, mig, t_max)
}

  ## Calculate the average to carrying capacity
  avg_gen_time <- mean(table_matrix)
  table_avg_c[q, 1] <- mig
  table_avg_c[q,2] <- avg_gen_time
  
  q <- q+1
  
  }
  print(table_avg_a)
  print(table_avg_b)
  print(table_avg_c)

# Plot the results
  plot(table_avg_a[,1], table_avg_a[,2], type= 'l', col= "blue", ylim= c(0, max(table_avg_a[,2], table_avg_b[,2], table_avg_c[,2])), xlab = "Mean number of migrants", ylab= "Mean number of mutant in carrying capacity", main = "Comparison of Decay Rates")

# Add the line for decay_rate = 0.5
  
  lines(table_avg_b[,1], table_avg_b[,2], col= "red")
  lines(table_avg_c[,1], table_avg_c[,2], col= "purple")
 

```
### What could we measure?

-   setting the mutation rate as varies,we try to test the change of time to carrying capacity with the change of the mutation rate

``` {r}

# Initial parameters
N_init_a <- 1000
N_init_A <- 0
carrying_capacity <- 1000
decay_rate <- 0.1
sel_coeff <- 0.3
mut_rate <- seq(0.001, 0.03, by = 0.001)  # Sequence of mutation rates to test
mig_rate_a <- 3
t_max <- 1000
n_reps <- 1000

# Matrix to store mutation rate and average generation time to carrying capacity
table_avg <- matrix(nrow = length(mut_rate), ncol = 2)

q <- 1  # Counter for each mutation rate value

# Outer loop: iterate over each mutation rate
for (mut in mut_rate) {

  # Vector to store generation times for each repetition
  table_matrix <- numeric(n_reps)

  # Inner loop: run `n_reps` simulations for each mutation rate
  for (rep in 1:n_reps) {
    # Run the simulation with the current mutation rate `mut`
    table_matrix[rep] <- simulate_pop_mig_logistic(N_init_a, N_init_A, carrying_capacity, decay_rate, sel_coeff, mut, mig_rate_a, t_max)
  }

  # Calculate and store the average generation time for the current mutation rate
  avg_gen_time <- mean(table_matrix)
  table_avg[q, 1] <- mut             # Store the current mutation rate
  table_avg[q, 2] <- avg_gen_time     # Store the average time to carrying capacity
  
  q <- q + 1  # Move to the next row in `table_avg`
}

# Print the table showing mutation rates and their corresponding avg generation times
print(table_avg)

# Plot mutation rate vs. average generation time to carrying capacity
plot(table_avg[, 1], table_avg[, 2], type = 'l', ylim = c(0, max(table_avg[, 2])),
     xlab = "Mutation Rate", ylab = "Average Generation Time to Carrying Capacity",
     main = "Effect of Mutation Rate on Generation Time to Carrying Capacity")


```

### What could we measure?

-   setting the pollution rate (decay rate)as varies,we try to test the change of time to carrying capacity with the change of the pollution rate

```{r}

# Initial parameters
N_init_a <- 1000
N_init_A <- 0
carrying_capacity <- 1000
pollution_rate <- seq(0.1, 0.3, by = 0.01)  # Sequence of pollution rates to test
sel_coeff <- 0.3
mut_rate <- 0.001
mig_rate_a <- 3
t_max <- 1000
n_reps <- 1000

# Matrix to store results: each row corresponds to a pollution rate and avg generation time
table_avg <- matrix(nrow = length(pollution_rate), ncol = 2)

q <- 1  # Row counter for each pollution rate

# Outer loop: iterate over each pollution rate
for (poll in pollution_rate) {

  # Initialize a vector to store generation times for each repetition
  table_matrix <- numeric(n_reps)

  # Inner loop: run `n_reps` simulations for each pollution rate value
  for (rep in 1:n_reps) {
    # Run the simulation with the current pollution rate `poll`
    table_matrix[rep] <- simulate_pop_mig_logistic(N_init_a, N_init_A, carrying_capacity, poll, sel_coeff, mut_rate, mig_rate_a, t_max)
  }

  # Calculate and store the average generation time for the current pollution rate
  avg_gen_time <- mean(table_matrix)
  table_avg[q, 1] <- poll              # Store the current pollution rate
  table_avg[q, 2] <- avg_gen_time       # Store the average time to carrying capacity
  
  q <- q + 1  # Move to the next row in `table_avg`
}

# Print the table showing pollution rates and their corresponding avg generation times
print(table_avg)

# Plot pollution rate vs. average generation time to carrying capacity
plot(table_avg[, 1], table_avg[, 2], type = 'l', ylim = c(0, max(table_avg[, 2])),
     xlab = "Pollution Rate", ylab = "Average Generation Time to Carrying Capacity",
     main = "Effect of Pollution Rate on Generation Time to Carrying Capacity")

```

## Analysis and Observations

-   **Impact of Migration Rate**\
    Increasing migration rates generally reduced the time needed to reach carrying capacity. However, a saturation point was identified, beyond which further increases in migration had minimal impact.

-   **Influence of Decay Rate**

    -   **Low Decay Rate (e.g., 0.1)**: Carrying capacity was reached more quickly due to minimal environmental stress.

    -   **High Decay Rate (e.g., 0.2)**: Higher decay rates delayed the attainment of carrying capacity, requiring greater migration levels to maintain population stability.

-   **Interaction Between Migration and Decay**\
    Migration served as a buffer against high decay rates, enabling the population to reach carrying capacity despite environmental stress. However, this effect diminished under extreme conditions.

-   **Key Biological Insights**

    -   **Buffering Effect**: While migration can offset the impacts of decay, it is not sufficient for severe environmental conditions.

    -   **Conservation Implications**: Enhancing connectivity in polluted environments can aid recovery, but reducing pollution remains essential.
